---
title: "Practical Machine Learning Project"
author: "Arturo Vera"
date: "11/16/2020"
output: html_document
---

# Prediction Assignment Writeup

## I. Overview

This project looks to predict the manner in which 6 participants performed some exercise as described below. This is the “classe” variable in the training set. The machine learning algorithm is applied to the 20 test cases available in the test data. The report is the final part of the Coursera’s course Practical Machine Learning.

## II. Background

Using devices such as Jawbone Up, Nike FuelBand, and Fitbit it is now possible to collect a large amount of data about personal activity relatively inexpensively. These type of devices are part of the quantified self movement – a group of enthusiasts who take measurements about themselves regularly to improve their health, to find patterns in their behavior, or because they are tech geeks. One thing that people regularly do is quantify how much of a particular activity they do, but they rarely quantify how well they do it. In this project, your goal will be to use data from accelerometers on the belt, forearm, arm, and dumbell of 6 participants. They were asked to perform barbell lifts correctly and incorrectly in 5 different ways. More information is available from the website here: http://groupware.les.inf.puc-rio.br/har (see the section on the Weight Lifting Exercise Dataset).

## III. Data Loading and Exploratory Analysis

### III.a Overview

The training data for this project are available here:

https://d396qusza40orc.cloudfront.net/predmachlearn/pml-training.csv

The test data are available here:

https://d396qusza40orc.cloudfront.net/predmachlearn/pml-testing.csv

### III.b Enviroment

```{r Enviroment, message = FALSE}
rm(list=ls()) 
setwd("~/Documents/Trabajo/Telmex/Coursera/Practical Machine Learning")
library(knitr)
library(caret)
library(rpart)
library(rpart.plot)
library(rattle)
library(randomForest)
library(corrplot)
set.seed(1)
```

### III.c Data Loading and Cleaning

Data is loading fron the url and the training dataset is partinioned to create a Training set (70% of the data) for the modeling process and a Test set (with the remaining 30%) for the validations.

```{r, Load Data}
# download the datasets
training <- read.csv(url("https://d396qusza40orc.cloudfront.net/predmachlearn/pml-training.csv"))
testing  <- read.csv(url("https://d396qusza40orc.cloudfront.net/predmachlearn/pml-testing.csv"))

# create a partition with the training dataset 
inTrain  <- createDataPartition(training$classe, p=0.7, list=FALSE)
TrainSet <- training[inTrain, ]
TestSet  <- training[-inTrain, ]
dim(TrainSet)
dim(TestSet)

# remove columns that contain NA missing values
TrainSet <- TrainSet[, colSums(is.na(TrainSet)) == 0] 
TestSet <- TestSet[, colSums(is.na(TestSet)) == 0] 

# remove variables with Nearly Zero Variance
NZV <- nearZeroVar(TrainSet)
TrainSet <- TrainSet[, -NZV]
TestSet  <- TestSet[, -NZV]

# remove identification only variables (columns 1 to 5)
TrainSet <- TrainSet[, -(1:5)]
TestSet  <- TestSet[, -(1:5)]
dim(TrainSet)
dim(TestSet)

# classe colum as factor
TestSet$classe <- as.factor(TestSet$classe)
```

With the cleaning process above, the number of variables for the analysis has been reduced to 54 only.

### III.d Correlation Analysis

A correlation among variables is analysed before proceeding to the modeling procedures.

```{r}
corMatrix <- cor(TrainSet[, -54])
corrplot(corMatrix, order = "FPC", method = "color", type = "lower", tl.cex = 0.8, tl.col = rgb(0, 0, 0))
```

## IV. Prediction Model Building

Three methods will be applied to model the regressions (in the Train dataset) and the best one (with higher accuracy when applied to the Test dataset) will be used for the quiz predictions. The methods are: Random Forests, Decision Tree and Generalized Boosted Model, as described below.
A Confusion Matrix is plotted at the end of each analysis to better visualize the accuracy of the models.

### IV.a Random Forest

```{r, Random Forest}
# model fit
controlRF <- trainControl(method="cv", number=3, verboseIter=FALSE)
modFitRandForest <- train(classe ~ ., data=TrainSet, method="rf", trControl=controlRF)
modFitRandForest$finalModel
plot(modFitRandForest)

# prediction on Test dataset
predictRandForest <- predict(modFitRandForest, newdata=TestSet)
confMatRandForest <- confusionMatrix(predictRandForest, TestSet$classe)
confMatRandForest

# plot model and matrix results
plot(confMatRandForest$table, col = confMatRandForest$byClass, main = paste("Random Forest - Accuracy =", round(confMatRandForest$overall['Accuracy'], 4)))
```

### IV.b Decision Trees

```{r, Decision Trees}
# model fit
modFitDecTree <- rpart(classe ~ ., data=TrainSet, method="class")
fancyRpartPlot(modFitDecTree)

# prediction on Test dataset
predictDecTree <- predict(modFitDecTree, newdata=TestSet, type="class")
confMatDecTree <- confusionMatrix(predictDecTree, TestSet$classe)
confMatDecTree

# plot matrix results
plot(modFitRandForest)
plot(confMatDecTree$table, col = confMatDecTree$byClass, main = paste("Decision Tree - Accuracy =", round(confMatDecTree$overall['Accuracy'], 4)))
```

### IV.c Generalized Boosted Model

```{r, Generalized Boosted Model}
# model fit
controlGBM <- trainControl(method = "repeatedcv", number = 5, repeats = 1)
modFitGBM  <- train(classe ~ ., data=TrainSet, method = "gbm", trControl = controlGBM, verbose = FALSE)
modFitGBM$finalModel
plot(modFitGBM)

# prediction on Test dataset
predictGBM <- predict(modFitGBM, newdata=TestSet)
confMatGBM <- confusionMatrix(predictGBM, TestSet$classe)
confMatGBM

# plot matrix results
plot(confMatGBM$table, col = confMatGBM$byClass, main = paste("GBM - Accuracy =", round(confMatGBM$overall['Accuracy'], 4)))
```

## Select and apply model to the Test Data

The accuracy of the 3 regression modeling methods above are:

Random Forest : 0.9975
Decision Tree : 0.7251
GBM : 0.9864
In that case, the Random Forest model will be applied to predict the 20 quiz results (testing dataset) as shown below.

```{r, Model}
predictTEST <- predict(modFitRandForest, newdata=testing)
predictTEST
```

